\textit{MSF: Modulated Sub-graph Finder}

### 2018-03-28 MSF_Paper_FA.tex
Update and Pseudocode

To overcome this limitation, we developed a novel approach to identify
concertedly modulated sub-graphs in the global cell signaling network,
based on the DGEA results of all genes tested. Thereby, expression
patterns of genes are integrated according to the topology of their
interactions and allow potentially to read the flow of information
from the perturbation source to the effectors. The described software,
names \texttt{Modulated Sub-graph Finder} is freely available at
\url{https://github.com/MariamFarman/Modulated-SubGraph-Finder}.

@@

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a three steps heuristic as described in the
following. The proceeding identification of modulated sub-graphs from
a network by \texttt{MSF} is presented graphical and as pseudocode in
Fig.~\ref{fig:algorithm}. \TODO{add pseudocode or better replace figure with pseudocode}

\begin{figure}[ht!]
		\centering
	\fbox{
		\begin{minipage}{8 cm}

	\begin{subfigure}{.6\textwidth}
		\centering
		\includegraphics[width=6.5cm]{Tgfb1.png}
		\caption{Input}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=6.5cm]{TGFBstep1.png}
		\caption{Step 1}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=6.5cm]{TGFBstep2.png}
		\caption{Step 2}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=6.5cm]{TGFBstep3.png}
		\caption{Step 3}
	\end{subfigure}
	\caption{Overview of \texttt{MSF} process steps exemplified on a simplified toy
          example. (a) Input:
          showing a network of genes with green nodes as genes with
          \textit{p}-values $< $ 0.05, and blue nodes are genes with
          \textit{p}-values $>$ 0.05, (b) Step 1: genes circled red
          show the two initial modulated sub-graphs identified,
          sub-graph 1 with genes M,K,C and sub-graph 2 with genes
          S,V,T, (c) Step 2: shows the modulated sub-graph 1 being
          extended to M,K,C,A,D while no extension in sub-graph 2 (d)
          Step 3: shows both the modulated sub-graphs merging together
          with the addition of 3 genes Z,W,P.\label{fig:algorithm}}
			\end{minipage}
}

\end{figure}

@@

To demonstrate its usefulness \texttt{MSF} is applied to RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to
the Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host
innate and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now there are no approved antiviral
drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}.
The initial targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. The
Ebola viral protein VP35 targets the host type~I interferon
IFN-$\alpha / \beta$ to block the early antiviral immune response
~\cite{Prins,Konde,Falasca,Cardenas,Olejnik}. It has been proposed
that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. This data was chosen for analysis because it was the most recent EBOV infection data and the infection was carried out in human macrophages, which would help see the innate immune response of host. The aim of
testing this data with \texttt{MSF} was to identify the
modulated sub-graphs and check if \texttt{MSF} is able to identify the
IFN-$\alpha / \beta$ gene as one of the sources.

EBOV infection count data was downloaded from GEO
(GSE84188). Differential gene expression analysis was performed on the
count data with edgeR package (version 3.4.2)~\cite{edgeR}. The DEG
analysis results generated by edgeR were used as input for \texttt{MSF}. The
EBOV infection experiments describe the course of infection at three time-points 6, 24 and 48 hour post
infection (hpi). For the earliest time point at 6 hpi, five modulated sub-graphs were
identified with 41, 107, and 18 genes. Two of the five sub-graphs were less than 4 genes long. Most of the genes part of the sub-graphs were cytokines, chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as
a subset of CD4+ T helper cells. IFNB1 and IFNA1 were both identified
as two of the possible sources. Most of the sources identified by
\texttt{MSF} were type~I interferon induced genes. At 24 hpi nine
modulated sub-graphs were identified, IFNA1 was identified as one of
the sources. For the last time-point again IFNB1 and IFNA1 were
identified as the two sources out of the possible sources from seven
modulated sub-graphs identified.

@@

As stated earlier IFN-$\alpha / \beta$ could be one of the target
genes of Ebola infection. We were able to successfully identify IFNA1
as a source in all three time-points and INFB1 in two of the
time-points. Although IFNA1 and IFNB1 were two of the most significant
gene in the later time points, \texttt{MSF} was able to detect them as
a source in the early time-point when the genes were not significant
based on the individual DGEA alone. Identifying these sources will
reduce the search space for potential target genes and can help the
biologist as the starting point of clinical testing for drugs and
vaccines against an infection.

@@

 Three main modulated sub-graphs identified by \texttt{MSF} at 6 hpi
 are shown in Fig.~\ref{fig:Sub-graph6hpi}. The graph shows different colored nodes as
 genes associated to different KEGG pathways and how they cross talk. The KEGG
 association and visualization on the modulated sub-graphs was done
 using StringApp~\cite{StringApp} in Cytoscape~\cite{Cyto}. The graph
 unveils how more than one stimulus coincides to prevent the
 progression of infection. Sub-graph 1 shows toll-like receptor,
 cytokine, chemokine and jak-stat genes leading to TNF and eventually
 going into apoptosis. Sub-graph 2 shows Extra-cellular matrix (ECM)
 receptor interaction genes which are suggested to be in interaction
 with Ebola glycoprotein (GP)~\cite{Veljkovic}. ECM receptor, Focal
 adhesion and P13k-Akt signaling genes are the sources from one end of
 the sub-graph. P13k-Akt signaling goes to FoxO signaling which
 eventually leads to cell cycle, which at later time-points is
 disrupted. From the other end of sub-graph 2 toll-like recpetor
 combined with cytosolic DNA sensing and cytokine-chemokine signaling
 activates Jak-stat which leads to the start of apoptosis with a
 activation of a few apoptosis genes. Cytokine is also seen to sink
 into natural killer mediated cytotoxicity. Sub-graph 3 shows how some
 of the cytokine and chemokine genes activate the natural killer
 mediated cytotoxicity and apoptosis.

@@

 To test \texttt{MSF's} correctness simulation was also conducted. For each time-point, one \texttt{MSF} identified sub-graph was taken. A random graph was generated with the same number of genes as the sub-graphs using the same interaction file and the identified sources were counted. The objective
 was to find less sources in the simulated sub-graphs than observed in the real sub-graphs identified by
 \texttt{MSF}. Simulation results showed a chance of $<$10$\%$
 of the 1000 random sub-graphs generated to get more sources than observed.

@@

Likewise we compared gene enrichment results of sub-graph genes
identified by \texttt{MSF} using Reactome FI to the Reactome
predefined pathway enrichment analysis. Although \texttt{MSF} and
Reactome showed good agreement on the dis-regulated pathways e.g
Signaling by Interleukins, Interleukin-10 signaling, Cytokine
Signaling in Immune system, Chemokine receptors bind chemokines,
RIG-I/MDA5 mediated induction of IFN-alpha/beta pathways, NF-kB
activation, Reactome as well could not identify Toll-like receptor
signaling. This shows that
\texttt{MSF} is able to identify the important modulated genes in the
sub-graphs at the earliest time-point of infection. A summary of the comparison between the different tools is given in the supplement Table 1.

@@

>>>>

To overcome this limitation, we developed a novel approach to identify
concertedly modulated sub-graphs in the global cell signaling network,
based on the DGEA results of all genes tested. Thereby, expression
patterns of genes are integrated according to the topology of their
interactions and allow potentially to read the flow of information
from the perturbation source to the effectors. The described software,
named \texttt{Modulated Sub-graph Finder} is freely available at
\url{https://github.com/MariamFarman/Modulated-SubGraph-Finder}.

@@

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a four steps heuristic as described in the
following. The proceeding identification of modulated sub-graphs from
a network by \texttt{MSF} is presented as pseudocode in flowchart in
Fig.~\ref{fig:pseudocodemsf}.

@@

To demonstrate its usefulness \texttt{MSF} is applied to RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to
the Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host
innate and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now there are no approved antiviral
drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}.
The initial targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. The
Ebola viral protein VP35 targets the host type~I interferon
IFN-$\alpha / \beta$ to block the early antiviral immune response
~\cite{Prins,Konde,Falasca,Cardenas,Olejnik}. It has been proposed
that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. This data was chosen for analysis because it was the most recent EBOV infection data and the infection was carried out in human macrophages, which would help us comprehend the innate immune response of host. The aim of
testing this data with \texttt{MSF} was to identify the
modulated sub-graphs and check if \texttt{MSF} is able to identify the
IFN-$\alpha / \beta$ gene as one of the sources.

EBOV infection count data was downloaded from GEO
(GSE84188). Differential gene expression analysis was performed on the
count data with edgeR package (version 3.4.2)~\cite{edgeR}. The DEG
analysis results generated by edgeR were used as input for \texttt{MSF}. The
EBOV infection experiments describe the course of infection at three time-points 6, 24 and 48 hour post
infection (hpi). For the earliest time point at 6 hpi, five modulated sub-graphs were
identified with 41, 107, and 18 genes. Two of the five sub-graphs were less than 4 genes long. Most of the genes part of the sub-graphs were cytokines, chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as
a subset of CD4+ T helper cells. IFNB1 and IFNA1 were both identified
as two of the possible sources. Most of the sources identified by
\texttt{MSF} were type~I interferon induced genes. At 24 hpi nine
modulated sub-graphs were identified with four main sub-graphs consisting of 38, 85, 148 and 167 genes. IFNA1 was identified as one of
the sources in the most significantly modulated sub-graph. For the last time-point 48 hpi, eleven modulated sub-graphs were identified. Eight of the sub-graphs were less than seven genes and main sub-graphs had 39, 71 and 210 genes. IFNB1 and IFNA1 were
identified as the two sources out of the possible sources.

@@

As stated earlier IFN-$\alpha / \beta$ could be one of the target
genes of Ebola infection. We were able to successfully identify IFNA1
as a source in all three time-points and INFB1 in two of the
time-points. Although IFNA1 and IFNB1 were two of the most significant
gene in the later time points, \texttt{MSF} was able to detect them as
a source in the early time-point when the genes were not significant
based on the individual DGEA alone. Identifying the possible sources will
reduce the search space for potential target genes and can help the
biologist as the starting point of clinical testing for drugs and
vaccines against an infection.

@@

 Three main modulated sub-graphs identified by \texttt{MSF} at 6 hpi
 are shown in Fig.~\ref{fig:Sub-graph6hpi}. the figure shows two forms of graphs, one graph has different colored nodes as
 genes associated to different KEGG pathways and how they cross talk. The KEGG
 association and visualization on the modulated sub-graphs was done
 using StringApp~\cite{StringApp} in Cytoscape~\cite{Cyto}. StringApp can retrieve functional enrichment for Gene Ontology terms, KEGG Pathways, and protein domains for a user-specified network. The other graph like a flow chart,
 unveils how more than one stimulus coincides to prevent the
 progression of infection. Sub-graph 1 shows toll-like receptor,
 cytokine, chemokine and jak-stat genes leading to TNF and eventually
 going into apoptosis. Sub-graph 2 shows Extra-cellular matrix (ECM)
 receptor interaction genes which are suggested to be in interaction
 with Ebola glycoprotein (GP)~\cite{Veljkovic}. ECM receptor, Focal
 adhesion and P13k-Akt signaling genes are the sources from one end of
 the sub-graph. P13k-Akt signaling goes to FoxO signaling which
 eventually leads to cell cycle, which at later time-points is
 disrupted. From the other end of sub-graph 2 toll-like recpetor
 combined with cytosolic DNA sensing and cytokine-chemokine signaling
 activates Jak-stat which leads to the start of apoptosis with a
 activation of a few apoptosis genes. Cytokine is also seen to sink
 into natural killer mediated cytotoxicity. Sub-graph 3 shows how some
 of the cytokine and chemokine genes activate the natural killer
 mediated cytotoxicity and apoptosis.

@@

 To test \texttt{MSF's} correctness simulation was also conducted. For each time-point, one \texttt{MSF} identified sub-graph was selected and  using simulation random graph was generated with the same number of genes as the sub-graphs utilizing the same interaction file. The simulation identified sources were counted. The objective
 was to find less sources in the simulated sub-graphs than observed in the real sub-graphs identified by
 \texttt{MSF}. Simulation results showed a chance of $<$10$\%$
 of the 1000 random sub-graphs generated to get more sources than observed.

@@

Likewise we compared gene enrichment results of sub-graph genes
identified by \texttt{MSF} using Reactome FI to the Reactome
predefined pathway enrichment analysis. Although \texttt{MSF} and
Reactome showed good agreement on the dis-regulated pathways e.g
Signaling by Interleukins, Interleukin-10 signaling, Cytokine
Signaling in Immune system, Chemokine receptors bind chemokines,
RIG-I/MDA5 mediated induction of IFN-alpha/beta pathways, NF-kB
activation, Reactome as well could not identify Toll-like receptor
signaling. This shows that
\texttt{MSF} is able to identify the important modulated genes in the
sub-graphs at the earliest time-point of infection.

@@

\begin{figure*}
	\centering
	\fbox{
		\begin{minipage}{17 cm}
		\includegraphics[width=1.0\linewidth]{PseudoCodeMSF}
		\caption{Shows the algorithm of MSF with four steps to find the modulated sub-graphs and identify sources and sinks.}
		\label{fig:pseudocodemsf}
	\end{minipage}
}

\end{figure*}
